---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
Record linkage and deduplication of individual-level data, such as repeated spells in hospital, or recurrent cases of infection is a common task in epidemiological analysis and other fields of research. 

The `diyar` package aims to provide a simple and flexible implementation of deterministic record linkage, and episode grouping for the application of case definitions in epidemiological analysis.

## Installation
```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("OlisaNsonwu/diyar")
```

## Usage
There are two main aspects of the `diyar` package; record and episode grouping. Additionally, `number_line` objects are used in both as representations of a range of values to match, and time periods respectively. 

### Number line objects
Series of real numbers on a number line. `diyar` also includes functions used to manipulate these objects. Some useful ones are shown below.
```{r number_line, message=FALSE, warning=FALSE}
library(diyar)
library(dplyr)

l <- as.Date("01/04/2019", "%d/%m/%Y")
r <- as.Date("30/04/2019", "%d/%m/%Y")
nl <- number_line(l, r)
nl
reverse_number_line(nl)
shift_number_line(nl, -2)
expand_number_line(nl, 2)
number_line_sequence(nl, by =3)
```

### Episode grouping
Group records into chronological episodes for the purpose of record deduplication and implementing case definitions in epidemiological analysis.

`fixed_episode()` and `rolling_episode()` are the simplest implementation this process in `diyar`.
```{r}
data(infections); infections
db_a <- infections
# Fixed episodes
fixed_episodes(x = db_a$date, case_length = db_a$epi_len, display = FALSE)
# Rolling episodes
rolling_episodes(x = db_a$date, case_length = db_a$epi_len, recurrence_length = 40, display = FALSE)

db_b <- mutate(db_a, epid_interval= fixed_episodes(x = date, case_length = epi_len, strata = infection, display = FALSE))
db_b$epid <- db_b$epid_interval@gid
db_b
```

`episode_group()` is a more comprehensive options and provides useful information on each episode.
```{r}
db_c <- episode_group(db_a, sn=rd_id, date = date, strata = infection, case_length = epi_len, display = FALSE, group_stats = TRUE)
db_c
```

### Record grouping
Multi-stage deterministic linkages that addresses missing values by using a specfied list of alternative matching criteria.
```{r}
# Two or more stages of record grouping
data(staff_records); staff_records
pids <- record_group(staff_records, sn = r_id, criteria = c(forename, surname),
                     data_source = sex, display = FALSE)
left_join(staff_records, pids, by=c("r_id"="sn"))
staff_records

# Range matching
dob <- select(staff_records, sex)
dob$age <- c(10,8,20,5,5,9,7)

# age range - age + 20 years
dob$range <- number_line(dob$age, dob$age+20, gid=dob$age)
bind_cols(dob, record_group(dob, criteria = sex, sub_criteria = list(s1a="range"), display = FALSE))

# age range - age +- 20 years
dob$range <- number_line(dob$age-20, dob$age+20, gid=dob$age)
bind_cols(dob, record_group(dob, criteria = sex, sub_criteria = list(s1a="range"), display = FALSE))
```

Find out more [here](https://olisansonwu.github.io/diyar/index.html)!.

