---
title: "Implementing case definitions for epidemiological analysis in R"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementing case definitions for epidemiological analysis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#Introduction
Deduplicating events is a major part of establishing a case definition in epidemiological analyses. In such analyses, you select a reference event (`"Case (C)"`) which is taken as the start of an episode. Subsequent events within a specified period after this reference event are then considered duplicates `"(D)"`.

`fixed_episodes()`, `rolling_episodes()` and `episode_group()` aim to provide a simple but flexible way of grouping records from multiple data sources into episodes. This then allows for easy deduplication or a sub-analysis within each episode.

#Uses
The flexible use of [`record_group`](record_group.html) and the episode grouping functions allows you to apply different case definitions to a dataset. Some examples include;

1. Create episodes of events that occur at a point in time e.g. medical diagnoses or traffic incidents
2. Create episodes of overlapping periods of events e.g. hospital stay. See [interval grouping](#interval_grouping)
3. Create episodes that reoccur within a defined period (`recurrence_length`) from the last event
3. Specify which event (or period of event) is taken as the `"Case (C)"`. See [case assignment](#case_assignment)
4. Create episodes with a different `case_length` and/or `recurrence_length` for different subsets of a dataset. See [stratified grouping](#stratified_grouping)

#Implementation 

<a id= "episode_grouping"></a>

## Overview
An episode as produced by `fixed_episodes()`, `rolling_episodes()` or `episode_group()` is a set of events (or periods of events) within a specific period of time. Each begins with a reference event - the `"Case (C)"`, and may contain a `"Recurrent (R)"` event. The remaining events are taken as duplicates `"(D)"`.

Two type of episodes can be produced;

- `"fixed"` episodes - where `"Duplicate (D)"` events (or periods of events) are those within a specified period (`case_length`) after/before the `"Case (C)"`. Any event after this `case_length` is grouped into a different episode.

- `"rolling"` episodes - where `"Duplicate (D)"` events (or periods of events) are those within the `case_length` and specified period recurrence. See [episodes windows and recurrence periods](#recurrence).

`"fixed"` and `"rolling"` episodes are created with `fixed_episdoes()` and `rolling_episodes()` respectively, while `episode_group()` creates both.

<a id ="recurrence" ><a/> 

##Episode windows and recurrence periods
An episode window (`epid_interval`) is the period between the `"Case"` and the last event in the episode (or the end point of the last period). A `"fixed"` episode is never longer than its `"case_length"`.

A recurrence period (`recurrence_length`) is the period after the `case_length` within which  repeat events are taken as a recurrence of the same episode. The first new record within this period is taken as a `"Recurrent (R)"` record while the rest are taken as duplicates `"(D)"` of that recurrent `"(R)"` record. This means that a rolling episode will continue to expand as far as there's a record within the recurrence period of the last event in the episode. The last event could be a `"Case (C)"`, `"Duplicate (C)"` or `"Recurrent (R)"` record. As a result, rolling episodes can be longer than their `case_length`.

The examples below demonstrates the difference;

***NOTE; `to_s4` and `to_s4()` changes these functions' output from a data.frame (current default) to `epid` objects. `epid` objects will be the default output in the next release.***

```{r message=FALSE, warning=FALSE}
library(dplyr); library(lubridate); library(diyar)

data("infections_2")
dates <- infections_2$date

# Fixed episodes
f <- fixed_episodes(date = dates, case_length=14, group_stats = TRUE, to_s4 =TRUE)

# Rolling episodes
r <- rolling_episodes(date = dates, case_length=14, group_stats = TRUE, display = FALSE, to_s4 =TRUE)

dates # dates

f # fixed episode identifiers

r # rolling episode identifiers
```

In the `"fixed"` episodes example, records 1 to 3 are assigned a unique episode ID (`"E-1"`). In this group, record 1 is the `"Case (C)"` and records 2 and 3 the duplicates `"(D)"`. This is because records 2 and 3 are within 15 days (difference of 14 days) of record 1. Record 4 occurred after this 15-day period and so begins a new `"Case (C)"` which is assigned a new episode ID (`"E-4"`). Record 5 is within 15 days of record 4, so is considered a `"Duplicate (D)"` of record 4 and assigned the same episode ID as record 4 (`"E-4"`). This process continues chronologically until all records have been assigned an episode ID.

In the `"rolling"` episodes example, records 1 to 3 are also grouped together as episode `"E-1"` however, unlike the `"fixed"` episode example, record 4 is not considered a new `"Case (C)"`. Instead, its assigned to episode `"E-1"` as a `"Recurrent (R)"` record. This occurred because it was within 15 days (difference of 14 days) of the last `"Duplicate (D)"` (record 3). Record 5 is also assigned to episode `"E-1"` because it's within 15 days of record 4. Record 6 is not within 15 days of the initial `"Case (C)"` (record 1) and not within 15 days of the last record at this stage (record 5), so it's now considered a new `"Case (C)"` and assigned a new episode ID (`"E-6"`).

\newline

**Figure 1: `"Fixed"` and `"rolling"` episodes with their respective `case_lenth` and `recurrence_length`**

```{r echo=FALSE, fig.height=4, fig.width=7, message=FALSE, warning=FALSE}
library(tidyr); library(ggplot2)

r_epids <- f_epids <- infections_2[c("rid","pid","date")]

f_epids$epids <- f
r_epids$epids <- r

r_epids$dt_a <- start_point(r_epids$epids@epid_interval)
r_epids$dt_z <- end_point(r_epids$epids@epid_interval)
r_epids$case_nm <- r_epids$epids@case_nm
r_epids$epid <- r_epids$epids@.Data

f_epids$dt_a <- start_point(f_epids$epids@epid_interval)
f_epids$dt_z <- end_point(f_epids$epids@epid_interval)
f_epids$case_nm <- f_epids$epids@case_nm
f_epids$epid <- f_epids$epids@.Data

plot <- bind_rows(
  mutate(r_epids, method="rolling episodes"),
  mutate(f_epids, method="fixed episodes")) %>% 
  mutate(epi=14, recur=14) %>% 
  select(method, rid, epid, date, dt_a, dt_z, epi, recur, case_nm) %>% 
  mutate(epid = paste("EP",epid,sep="")) %>% 
  mutate_at(vars(dt_a, dt_z), funs(dmy(format(.,"%d/%m/%Y")))) 

plot <- plot %>% 
  group_by(method, epid) %>% 
  arrange(date) %>% 
  mutate(
    last_record = ifelse(row_number()==n(),1,0),
    recur_dup = ifelse(case_nm=="Duplicate" & !is.na(case_nm[row_number()+1]) & case_nm[row_number()+1] == "Recurrent", 1,0)
    ) %>% 
  ungroup() %>% 
  mutate(
    dash_range = case_when(
      case_nm=="Case" ~ epi, 
      (case_nm =="Recurrent" & last_record !=1) | (recur_dup ==1) ~ recur,
      TRUE ~ 0
    ), 
    dash_num = ifelse(dash_range==0,"",paste(dash_range, "days")),
    dash_range = date + duration(dash_range,"days")
  )

g_area <- unique(select(plot, method, epid, date=dt_a, dt_z)) 
g_area$rid <- max(plot$rid)

ggplot(plot, aes(x=date, y=rid, group=rid, colour=epid)) +
  geom_point() +
  geom_segment(aes(x=date,xend=dash_range,y=rid,yend=rid), linetype=2) +
  geom_text(aes(hjust="right", label=case_nm), nudge_x = -2, nudge_y = .05, show.legend = FALSE) +
  geom_text(aes(hjust="left", label=dash_num), nudge_x = 2, nudge_y = .18, show.legend = FALSE, size =2.5) +
  scale_y_continuous("Record ID", breaks = 1:6) +
  scale_x_date("Specimen date", limits = c(dmy("15/03/2019"), dmy("01/07/2019") ), date_breaks = "25 days", date_labels = "%d %b") +
  facet_wrap(~method, scales = "free_y") +
  guides(col = guide_legend(nrow = 1)) +
  theme(
    strip.background = element_rect(fill="transparent"),
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

***If your case definition does not explicitly require a rolling episode, use `fixed_episode()`. It takes less time to complete.***

<a id="case_assignment"></a> 

## Case assignment
This section covers the different ways to choose which record is taken as the `"Case (C)"`.

### Number of recurrence periods and episodes
You can choose to have a certain number of episodes per `strata` by using `episode_max`. When `episode_max` is reached, any record not yet grouped is assigned a unique episode ID, making them unique cases `"(C)"`.

```{r warning=FALSE, message=FALSE}
data("infections_3");
dbs <- infections_3[c("pid","date")]; dbs

# Maximum of one fixed episode grouping per strata
dbs$eps_1 <- fixed_episodes(strata = dbs$pid, date =dbs$date, case_length = 3, display = FALSE, episodes_max = 1, to_s4 = TRUE)

# Maximum of one rolling episode grouping per strata
dbs$eps_2 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 3, display = FALSE, episodes_max = 1, to_s4 = TRUE)

# Maximum of two fixed episode grouping per strata
dbs$eps_3 <- fixed_episodes(strata = dbs$pid, date =dbs$date, case_length = 3, display = FALSE, episodes_max = 2, to_s4 = TRUE)

dbs

```

By default, `rolling_episodes()` will continue checking for `"Recurrent" (R)` records indefinitely, but you can limit the number of recurrence periods an episode has by using `rolls_max`.

```{r warning=FALSE, message=FALSE}
# Infinite recurrence periods per episode (Default)
dbs$eps_4 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = FALSE, to_s4 = TRUE)

# Maximum of one recurrence period per episode
dbs$eps_5 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = FALSE, 
                              rolls_max = 1, to_s4 = TRUE)

# Maximum of two recurrence periods per episode
dbs$eps_6 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = FALSE, 
                              rolls_max = 2, to_s4 = TRUE)

dbs
```

### Chronological order of events
By default, episode grouping begins at the earliest event (or period of event) and proceeds to the most recent one, making the earliest event the `"Case (C)"`. The opposite of this is to begin episode grouping at the most recent record and proceed backwards in time. As a result, the most recent record is taken as the `"Case (C)"`. To do this, change `from_last` to `TRUE`.

```{r message=FALSE, warning=FALSE}
dbs <- infections_2[c("date")]; dbs

# Episode grouping in chronological order
dbs$forward <- fixed_episodes(date=dbs$date, case_length= 14,
                              group_stats = TRUE, display = FALSE, to_s4=TRUE)

# Episode grouping in reverse chronological order
dbs$backward <- fixed_episodes(date=dbs$date, case_length= 14, group_stats = TRUE, display = FALSE,
                               from_last=TRUE, to_s4=TRUE)

dbs[c("forward","backward")]
```

<a id ="custom_sort"></a>

### User defined case assignment
You can use a specific preference for case assignment by using `custom_sort`. In `fixed_episodes()` and `rolling_episodes()`, this is a vector whose values when sorted in ascending order specifies the required preference. For example, records with a `custom_sort` value of `1` will be preferentially taken as the start of an episode over one with a value of `2`. This is prioritised over the chronological order of those records. See the example below.

```{r message=FALSE, warning=FALSE}
dates <- c("01/04/2019", "05/04/2019", "07/04/2019")
dates <- as.Date(dates,"%d/%m/%Y")
user_sort <- c(2,1,2)

# preference determined by from_last 
fixed_episodes(date=dates, case_length=6, to_s4=TRUE, display=FALSE, group_stats = TRUE)

# user defined preference via custom sort is prioritised before from_last 
fixed_episodes(date=dates, case_length=6, to_s4=TRUE, custom_sort = user_sort, display=FALSE, group_stats = TRUE)

# user defined preference via custom sort is prioritised before from_last. Duplicates flagged from both directions
fixed_episodes(date=dates, case_length=6, to_s4=TRUE, custom_sort = user_sort, display=FALSE, 
               bi_direction = TRUE, group_stats = TRUE)

```

In the second example above, even though the second record occurred after the first, episode grouping began at the second one and continued chronologically in the direction determined by `from_last`. 

A consequence of using `custom_sort` was that record 1 was not grouped together with record 2 even though its within the `case_length` of record 2. This is because duplicates are tracked in one direction which is also determined by `from_last`. To track duplicates from both directions use `bi_direction` as shown in the third example.   

Note that examples 1 and 3 now have the same number of duplicates (and episodes) but different cases - records 1 and 2 respectively.

***`custom_sort` will follow `R` standard sort behaviour. For example, a `factor` will sort on its levels not the descriptive label.***

For a practical example, we'll use this feature on a case definition where *E. coli* urinary tract infections (UTI) are considered precursors to *E. coli* bloodstream infections (BSI). This means that episodes need to be created in such a way that; if there are UTI and BSI records within the same `case_length`, the UTI record will be taken as the `"Case (C)"`. 

```{r message=FALSE, warning=FALSE}
dbs <- infections_2[c("date","infx")]; dbs
dbs$infx <- gsub("E. coli ","",dbs$infx)
dbs$infx[c(2,5)] <- "UTI"

dbs$epids_1 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx, display = FALSE, to_s4 = TRUE)

dbs$infx_f <- factor(dbs$infx, levels = c("UTI","BSI"))

dbs$epids_2 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx_f, display = FALSE, to_s4 = TRUE)

dbs$epids_3 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx_f, display = FALSE, to_s4 = TRUE, bi_direction = TRUE)

dbs
```

In `epids_2`, after changing the sort preference using factor levels, record 1 (*E. coli* BSI)  and record 2 (*E. coli* UTI) are no longer part of the same episode. This is because record 2, is now the reference record where episode grouping began, and since record 1 occurred before record 2, they are not grouped together. `epids_3` is episode grouping when `bi_direction` is used.

***In `episode_group()`, you can implement `custom_sort` in levels. You'll need to create a column for each level. The column names are then supplied to `custom_sort` in the required order (level)***

<a id="interval_grouping"></a>

## Interval grouping
In this section we discuss the process of grouping periods of events into episodes. Each period/interval is essentially a record with a start and end point in time. For the purpose of episode grouping, these periods are created as `number_line` objects and supplied to the `date` argument.

Below are simple examples;

```{r message=FALSE, warning=FALSE}
dbs <- tibble(date=c("01/04/2019", "05/04/2019"))

dbs$date <- as.Date(dbs$date, "%d/%M/%Y")
dbs$period <- number_line(dbs$date, dbs$date + 10)

dbs

# Grouping events
fixed_episodes(date=dbs$date, case_length=30, to_s4=TRUE, display=FALSE, group_stat=TRUE)

# Grouping periods
fixed_episodes(date=dbs$period, case_length=30, to_s4=TRUE, display=FALSE, group_stat=TRUE)
```

As a practical example, we will group periods of hospital stay into episodes. This is different from grouping the actual admission or discharge events.

```{r message=FALSE, warning=FALSE}
data("hospital_admissions"); hospital_admissions

hospital_admissions$admin_period <- number_line(hospital_admissions$admin_dt, hospital_admissions$discharge_dt)

# Grouping the actual admissions into episodes
fixed_episodes(date=hospital_admissions$admin_dt, sn=hospital_admissions$rd_id, case_length = 0, 
                display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Grouping the periods of stay (admission -> discharge)
fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
                display = FALSE, to_s4 = TRUE, group_stats = TRUE)

```

Periods are grouped into the same episode if they overlap. Since this can happen in different ways, you can choose how which methods (`overlap_method`) should be considered in your analysis. The options available are; `"across"`, `"inbetween"`, `"chain"`, `"aligns_start"` and `"aligns_end"`. The default option is to use all. Below are examples demonstrating each method.

```{r warning=FALSE}
# Overlapping intervals
across <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = "across", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

across

# Chained intervals
chain <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = "chain", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

chain

# Intervals with aligned end points
aligns_end <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = "aligns_end", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

aligns_end

# Intervals with aligned start points
aligns_start <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = "aligns_start", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

aligns_start

# Intervals occurring completely within others
inbetween <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = "inbetween", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

inbetween

# Chained intervals and those occurring completely within others
chain_inbetween <- fixed_episodes(date=hospital_admissions$admin_period, sn=hospital_admissions$rd_id, case_length = 0, 
               overlap_method = c("chain","inbetween"), display = FALSE, to_s4 = TRUE, group_stats = TRUE)

chain_inbetween

```

**Figure 2: Different options for `overlap_method` using a case_length of `"0"` days.**
```{r echo=FALSE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
across <- cbind(hospital_admissions, to_df(across))
chain <- cbind(hospital_admissions, to_df(chain))
aligns_start <- cbind(hospital_admissions, to_df(aligns_start))
chain_inbetween <- cbind(hospital_admissions, to_df(chain_inbetween))
inbetween <- cbind(hospital_admissions, to_df(inbetween))
aligns_end <- cbind(hospital_admissions, to_df(aligns_end))

plot <- rbind(
  mutate(across, method="across"),
  mutate(chain, method="chain"),
  mutate(aligns_start, method="aligns_start"),
  mutate(aligns_end, method="aligns_end"),
  mutate(inbetween, method="inbetween"), 
  mutate(chain_inbetween, method="chain and inbetween")
  )

plot <- mutate(plot,
  mrk=paste(epid, method,sep="_"),
  admin_dt = dmy(format(left_point(admin_period), "%d/%m/%Y")),
  discharge_dt = dmy(format(right_point(admin_period), "%d/%m/%Y"))
  ) %>% select(-admin_period)

plot <- dplyr::arrange(plot, .data$mrk)
plot$N <- rep(rle(plot$mrk)$lengths, rle(plot$mrk)$lengths)

plot <- plot %>%
  mutate(epid_2 = paste(method, ": epid ",epid, sep="" )) %>% 
  select(rd_id, epid, epid_2, admin_dt, discharge_dt, method, N, epid, case_nm) %>%
  filter(N>1) %>%
  select(rd_id, admin_dt, discharge_dt, method, epid, epid_2, case_nm) %>%
  gather(var, val, -c(rd_id, method, epid, epid_2, case_nm)) %>%
  mutate(
    rd_id = as.character(rd_id), 
    lab =ifelse(case_nm=="Case" & var=="admin_dt", case_nm, ""), 
    method = factor(method, levels = c("across","aligns_start","aligns_end","chain", "inbetween", "chain and inbetween"))
    )

ggplot(plot, aes(x=val, y=rd_id, group=rd_id, label=lab, colour=epid_2)) +
  geom_line(aes(colour=epid_2)) +
  geom_point() +
  geom_text(aes(hjust="right"), nudge_x = -2, nudge_y = .05, show.legend = FALSE) +
  scale_y_discrete("Record ID") +
  scale_x_date("Admitted period", limits = c(dmy("15/12/2018"), dmy("10/02/2019")), date_breaks = "20 days", date_labels = "%d %b") +
  guides(col = guide_legend(nrow = 2)) +
  facet_wrap(~method, scales = "free_y") +
  theme(
    strip.background = element_rect(fill="transparent"),
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

```

```{r warning=FALSE, include=FALSE}

# Overlapping intervals
ha2 <- hospital_admissions
ha2$epi_len <- 30
ha2$shift <- ifelse(ha2$rd_id %in% c(2,8,5,7), 0, 30)
ha2$admin_period <- shift_number_line(ha2$admin_period, duration(ha2$shift, "days"))

# Overlapping intervals
across <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = "across", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Chained intervals
chain <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = "chain", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Intervals with aligned end points
aligns_end <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = "aligns_end", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Intervals with aligned start points
aligns_start <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = "aligns_start", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Intervals occurring completely within others
inbetween <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = "inbetween", display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Chained intervals and those occurring completely within others
chain_inbetween <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = c("chain","inbetween"), display = FALSE, to_s4 = TRUE, group_stats = TRUE)

# Chained intervals and those with aligned end points
chain_aligns_end <- fixed_episodes(date=ha2$admin_period, sn=ha2$rd_id, case_length = 30, 
               overlap_method = c("chain","aligns_end"), display = FALSE, to_s4 = TRUE, group_stats = TRUE)

```

**Figure 3: Different options for `overlap_method` using a case_length of `"30"` days**
```{r echo=FALSE, fig.height=5.5, fig.width=8, message=FALSE, warning=FALSE}
across <- cbind(ha2, to_df(across))
chain <- cbind(ha2, to_df(chain))
aligns_start <- cbind(ha2, to_df(aligns_start))
chain_inbetween <- cbind(ha2, to_df(chain_inbetween))
chain_aligns_end <- cbind(ha2, to_df(chain_aligns_end))
inbetween <- cbind(ha2, to_df(inbetween))
aligns_end <- cbind(ha2, to_df(aligns_end))

plot <- rbind(
  mutate(across, method="across"),
  mutate(chain, method="chain"),
  mutate(aligns_start, method="aligns_start"),
  mutate(aligns_end, method="aligns_end"),
  mutate(inbetween, method="inbetween"), 
  mutate(chain_inbetween, method="chain and inbetween"), 
  mutate(chain_aligns_end, method="chain and aligns_end")
  )

plot <- mutate(plot,
  mrk=paste(epid, method,sep="_"),
  admin_dt = dmy(format(left_point(admin_period), "%d/%m/%Y")),
  discharge_dt = dmy(format(right_point(admin_period), "%d/%m/%Y"))
  ) %>% select(-admin_period)

plot <- dplyr::arrange(plot, .data$mrk)
plot$N <- rep(rle(plot$mrk)$lengths, rle(plot$mrk)$lengths)

plot <- plot %>%
  mutate(epid_2 = paste(method, ": epid ",epid, sep="" )) %>% 
  select(rd_id, epid, epid_2, admin_dt, discharge_dt, method, N, epid, case_nm, epi_len) %>%
  filter(N>1) %>%
  select(rd_id, admin_dt, discharge_dt, method, epid, epid_2, case_nm, epi_len) %>%
  gather(var, val, -c(rd_id, method, epid, epid_2, case_nm, epi_len)) %>%
  mutate(
    rd_id = as.character(rd_id), 
    lab =ifelse(case_nm=="Case" & var=="admin_dt", case_nm, ""), 
    method = factor(method, levels = c("across","aligns_start","aligns_end","chain", "inbetween",
                                       "chain and inbetween", "chain and aligns_end","aligns_start and aligns_end"))
    )

ep_plot <- filter(plot, var=="discharge_dt" & case_nm == "Case") %>% 
  mutate(var="start")

#ep_plot$k <- ifelse(ep_plot$case_nm=="Case", ep_plot$epi_len, ep_plot$recur_len)

ep_plot <- mutate(ep_plot, val= val + duration(epi_len,"days"), var="end") %>% 
  bind_rows(ep_plot)

ep_plot$lab <- ifelse(ep_plot$var=="start", paste(ep_plot$epi_len, "days"),"")
                      
ggplot(plot, aes(x=val, y=rd_id, group=rd_id, label=lab, colour=epid_2)) +
  geom_line() +
  geom_point() +
  geom_text(aes(hjust="right"), nudge_x = -2, nudge_y = .05, show.legend = FALSE) +
  geom_line(data=ep_plot, aes(x=val, y=rd_id, group=rd_id, colour=epid_2), linetype=2) +
  #geom_text(data=ep_plot, aes(hjust="left", label=lab), nudge_x = 2, nudge_y = .18, show.legend = FALSE, size =2.5) +
  scale_y_discrete("Record ID") +
  scale_x_date("Admitted period", limits = c(dmy("15/12/2018"), dmy("01/04/2019")), date_breaks = "35 days", date_labels = "%d %b") +
  guides(col = guide_legend(nrow = 4)) +
  facet_wrap(~method, scales = "free_y") +
  theme(
    strip.background = element_rect(fill="transparent"),
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

```


<a id="stratified_grouping"></a>

## Stratified episode grouping
Episode grouping can be done separately for different subsets (`strata`) of the dataset. Examples of a `strata` could be patient IDs, pathogen, source of infection or any combination of these. Episodes will be limited to each `strata` however, episodes from different `strata` can have different `case_length` and/or `recurrence_length`.

**`record_group()` is useful for creating group identifiers which can used as a `strata`. See [record group](record_group.html) for further details.**

As an example, using the `infections` dataset, a case definition may specify the following;

- UTI and BSI records are considered different episodes regardless of when they occur
- UTI has a `case_length` of 7 days and recurrence period of 30 days
- BSI has a `case_length` of 14 days and no recurrence period

In the example below, adding `infection` to the `strata` argument means that `"UTI"` and `"BSI"` records will not be in the same episode. Furthermore, the `case_length` and `recurrence_length` for each will be different as specified in the `epi` and `recur` columns respectively 
```{r warning=FALSE}
data(infections)
dbs <- infections[c("date","infection")]
dbs$epi <- ifelse(dbs$infection=="UTI", 7, 14)
dbs$recur <- ifelse(dbs$infection=="UTI", 30, 0)

dbs$epids <- rolling_episodes(date=dbs$date, case_length =dbs$epi, to_s4 =TRUE,
                            recurrence_length = dbs$recur, display = FALSE, group_stats = TRUE)

dbs
```

<a id="sub_strata"></a>

### Sub-strata
A sub-strata is created when records within a `strata` have a different `case_length` or `recurrence_length`. The case definition below demonstrates how this can be used.

- UTI and BSI records are different episodes regardless of when they occur. 
- UTI have a `case_length` of 7 days and recurrence period of 30 days
- BSI have a `case_length` of 14 days if not treated, **OR** `case_length` of 4 days if treated, and no recurrence period in both situations
- Respiratory tract infections (RTI) have a `case_length` of 28 days and recurrence period of 5 days

In this example, whether or not the infection is treated should be taken as the sub-strata, and the source of infection taken as the `strata`.

```{r warning=FALSE}
data("infections_4"); 
dbs <- infections_4

dbs$epids <- episode_group(infections_4, sn=rid, strata=c(pid, organism, source), date=date, 
                case_length =epi, episode_type = "rolling", recurrence_length = recur,
                display = FALSE, to_s4 = TRUE)

dbs
```

There are a few things to note with stratified grouping;

- Unless required, `case_length` and `recurrence_length` should be consistent across each `strata` otherwise, you'll inadvertently create a sub-strata
- Episode grouping with and without a sub-strata is different and could lead to different results
- Using sub-strata is not the same as adding that sub-strata to the `strata` argument. In the example above, adding `treated` to `strata` will group treated infections separately from untreated infections. While this could be the desired outcome depending on your case definition, the case definition above did not required treated and untreated infections to be grouped separately, only that the treated infections last longer.

Stratified grouping is the same as a separate analysis for each subset (`strata`) of the dataset.

<a id="tips"></a>

## Useful ways of using these function 
### Episode grouping across other units of time
In the examples above, episode grouping was done by days (`episode_unit`). However, it can be done by other units of time e.g. hours, days or weeks. Acceptable options are those supported by `lubridate's duration()` function. Below is an example of episode grouping by the hour.

```{r warning=FALSE, message=FALSE}
data("hourly_data"); dbs <- hourly_data

dbs[c("datetime", "category")]

rolling_episodes(strata = dbs$category, date = dbs$datetime, case_length = 5,
                 episode_unit = "hours", recurrence_length = 9, group_stats = TRUE, to_s4 = TRUE, display = FALSE)
```

### Limit episode grouping to a subset of the dataframe
For example, with the `hourly_data` dataset, you can decide to exclude `"GP1"` and `"GP2"` records from episode grouping as shown below. These records will be excluded from episode grouping and assigned a unique episode IDs.

```{r warning=FALSE, message=FALSE}
dbs <- head(hourly_data[c("datetime","category")], 10)
dbs$subset <- ifelse(dbs$category!="GP3", NA, "group")

dbs$epids <- rolling_episodes(strata= dbs$subset, date = dbs$datetime, case_length = 5, episode_unit = "hours", 
                        recurrence_length = 9, display = TRUE, group_stats = TRUE, to_s4 = TRUE)

dbs
```

### Use a `strata` from `record_group()`
You can used `pid` objects from `record_group()` as a `strata`.

```{r warning=FALSE, message=FALSE}
data(infections) 

dbs <- infections[c("date","infection")]; dbs

# familiar unique record ids for reference
rd_id <- c(640,17,58,21,130,79,45,300,40,13,31)

# strata based on matching sources of infection
dbs$pids <- record_group(dbs, criteria = infection, to_s4 = TRUE, display = FALSE)

# stratified grouping 
dbs$epids <- fixed_episodes(sn = rd_id, date = dbs$date, strata = dbs$pids, 
                             to_s4 = TRUE, display = FALSE, group_stats = TRUE, case_length = 10)

dbs
```

# Conclusion
There are a variety of ways to use these function. Before beginning, It's worth reviewing your case definition and its implication on the dataset. In general, the following steps will guide you on how to use them;

1. Work out which columns should be the `strata`
2. Choose whether you need `"fixed"` or `"rolling"` episodes
3. Choose whether you are grouping individual events or a duration of events by supply a date, datetime or `number_liine` objects as required. See [interval grouping](#interval_grouping)
4. Create a column for `case_length`, and/or `recurrence_length`. The values should be unique to each `strata` unless you require a [sub-strata](#sub_strata)
5. Change `from_last` to `TRUE` if you want to start episode grouping at the most recent record thereby making it the `"Case"`. Note that this is not the same as starting episode grouping at the earliest record (`from_last` is `FALSE`) and then picking the most recent record in that episode as the `"Case"`. See [case assignment](#case_assignment)
6. If you require the `"Case"` to be the earliest or most recent record of a particular type of record, use `custom_sort` in combination with `from_last`. If not, ignore this argument. See [user defined case assignment](#custom_sort)
7. If you require episodes to include records on either side of the `"Case"` use `bi_direction`. If not, ignore this argument
8. Determine if you are measuring episodes by the minute, hour or day etc., and set `episode_unit` accordingly. Be sure to supply a datetime to the `date` arguement if `episode_unit` is less than "days"
