---
title: "Implementing case definitions for epidemiological analysis in R"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementing case definitions for epidemiological analysis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
plot_epid <- diyar:::plot_epid
```
#Introduction
Deduplicating events is an important step in establishing a case definition in epidemiological analyses. For the most part, this involves selecting a reference event which represents an episode. Subsequent events within a specified period are then taken as duplicates of this reference event.

`fixed_episodes()`, `rolling_episodes()` and `episode_group()` provide simple but flexible ways of grouping records from multiple datasets into episodes. This then allows for easy deduplication or a sub-analysis within each episode.

#Uses
You can use [`record_group()`](record_group.html) and these episode grouping functions to implement a variety of case definitions. For example, they can group episodes;

+ from events that occur at points in time e.g. medical diagnoses or traffic incidents
+ from events that occur within periods of time e.g. hospital stay. See [interval grouping](#interval_grouping)
+ with periods of recurrence (`recurrence_length`)
+ with specific events as their reference event (case). See ["Making a case definition"](#case_assignment)
+ with different durations (`case_length`) or recurrence periods (`recurrence_length`) per subset (`strata`) of a dataset. See [stratified grouping](#stratified_grouping)

#Implementation 

<a id= "episode_grouping"></a>

## Overview
An episode as produced by `fixed_episodes()`, `rolling_episodes()` or `episode_group()` is a set of dated records (events or periods of events) within a specific period. Each episode begins with a reference event (case) and may contain duplicate and recurrent events. Duplicate event are those that occur between cases and recurrent.

These functions produce two main type of episodes - `"fixed"` episodes which do not include periods of recurrence and `"rolling"` episodes which do. `fixed_episodes()` and `rolling_episodes()` are wrapper functions of `episode_group()`.

<a id ="recurrence" ><a/> 

## Episodes, windows, case lengths and recurrence periods
Before we begin, let's go over some terms which will pop up now and again. An `event` is simply a dated record (single time point). A `period of event` or `event period` is a record with a start and end time points. The "lengths" - `case_length` and `recurrence_length` are specific durations after an `event`. More specifically, the `case_length` is a specified period after the `reference event` (first `event` of an `episode`) while `recurrence_length` is a specified duration after the other `events` subsequent to the `reference event`. 

All `events` within the same `case_length` or `recurrence_length` form a `window`. A `window` equates to an `episode`. `events` outside a `window` belong to separate `episodes`. What differentiates `rolling episodes` from `fixed episodes` is that when two or more `windows` overlap, they are collapsed into a larger `window`, indicating that the `episode` (first `window`) is still on-going up to the end of the second `window`. As a result, every `event` belonging to both `windows` are considered part of the same `"roliing episode"`. Consequently, a `"fixed episode"` is never longer than its `case_length` but a `"rolling episode"` becomes longer than its `case_length` if the first `window` (also referred to `case window`) overlaps with another.

The examples below demonstrate this difference with 4 events/records.

```{r message=FALSE, warning=FALSE}
library(diyar)

# Events
ds <- c("01/04/2019", "03/04/2019", "13/04/2019","16/04/2019")
ds <- data.frame(date= as.Date(ds, "%d/%m/%Y"))
ds$date
```

**Figure 1a: Illustration of key terms - fixed episodes**
```{r warning=FALSE}
# 6-day (5-day difference) episodes - fixed episodes
ds$f1 <- fixed_episodes(ds$date, case_length = 5, display = F)

ds$f1
```

```{r echo=FALSE, fig.height=4, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = ds$f1, date= ds$date, case_length = 5)
```

**Figure 1b: Illustration of key terms - rolling episodes**
```{r warning=FALSE}
# 6-day (5-day difference) episodes with 11-days recurrence periods - rolling episodes
ds$r1 <- rolling_episodes(ds$date, case_length = 5, recurrence_length = 10, display = F)

ds$r1
```

```{r echo=FALSE, fig.height=4, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = ds$r1, date= ds$date, case_length = 5, recurrence_length = 10)
```

### Let's go through another example ...
```{r message=FALSE, warning=FALSE}
dates <- diyar::infections_2$date

dates # dates
```

**Figure 2a: 15-day "fixed" episodes**

```{r message=FALSE, warning=FALSE}
# Fixed episodes
f <- fixed_episodes(date = dates, case_length=14, group_stats = T, to_s4 =T, display =F)

f # fixed episode identifiers
```


```{r echo=FALSE, fig.height=4, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = f, date= dates, case_length = 14)
```

In the `fixed episodes` example, records 1 to 3 are assigned to a `window` (i.e. window ID `1`). In this window, record 1 is the `reference event` (case) and records 2 and 3 are the  `duplicates`, because they are within 15 days (14-day difference) of record 1. This `window` then becomes one `episode` (episode ID `1`; Figure 2a). Record 4 is not within this 15-day period after record 3 and so begins a new `window` (and `episode` i.e. window and episode ID `4`). Record 5 is within 15 days of record 4 and so is assigned to the same window and episode (window and episode ID `4`). This process continues chronologically until all records have been assigned to an `episode`.

**Figure 2b: 15-day "rolling" episodes with a 15-day recurrence period**

```{r message=FALSE, warning=FALSE}
# Rolling episodes
r <- rolling_episodes(date = dates, case_length=14, group_stats = T, display = F, to_s4 =T)

r # rolling episode identifiers
```

```{r echo=FALSE, fig.height=4, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = r, date= dates, case_length = 14, recurrence_length =14)
```

In the `rolling episodes` example, records 1 to 3 are also assigned to one window (window ID `1`) however, unlike the `fixed episode` example, since records 4 and 5 both fall within 15 days (14-day difference) of the last `event` in the previous `window` (record 3; Figure 2b), they are both folded into `window` ID `1` to give a longer `episode` ID `1`. Since record 4 is the earliest of the two, it's flagged as the `recurrent event` while record 5 is taken as a `duplicate`. 

***If your case definition does not explicitly require a rolling episode, use `fixed_episode()`. It takes less time.***

For `rolling` episodes, it may be useful to imagine the `windows` as mini-`episodes` that may eventually be collapsed into one main `episode`. As mentioned, these mini-`episodes` will only be collapsed if they overlap. Each `window` is either a `case window` or `recurrence window` and will have one `event` flagged as the `case` or `recurrent event` respectively. 

`case` events are always `reference events` for `case windows` and the `episode` as a whole. Conversely, by default, `recurrent events` are not `reference events` for `recurrence windows`. Strictly speaking, the `reference event` for a `recurrence window` is the **last** `event` of the previous `window` and this is where the `recurrence_length` starts from. In the example above, that is record 3. 

<a id="case_assignment"></a> 

## Making a case definition
Almost every arguments in `fixed_episodes()`, `rolling_episodes()` and `episode_group()` can affect which `events` are taken as the `case` and `recurrent event`, and this ultimately allows you apply a variety of case definitions to your datasets. This section covers the main parts of this process and which of these you can control or change.  

### Reference events
From `diyar v0.1.0`, you can use the `recurrence_from_last` argument to specify that the `reference event` for a `recurrence window` should be the **first** `event` from the previous `window` and not the **last** (default). This means that after flagging a `recurrent event`, the `recurrence window` is re-calculated from it. This new option is only really useful if your `case_length` and `recurrence_length` are different. If they are the same (which is the default behaviour), you'll have essentially created a `fixed episode`. See a demonstration of this below.

```{r message=FALSE, warning=FALSE}
df <- c("01/04/2019", "11/04/2019", "16/04/2019","21/04/2019", "07/05/2019")
df <- data.frame(date= as.Date(df, "%d/%m/%Y"))
# 15-day fixed episodes
df$f1 <- fixed_episodes(df$date, case_length = 14, display = F)

# 15-day rolling episods with 15-day recurrence periods
  # Since the first event is the reference event for the next recurrence window and,
  # the recurrence_length is the same the case_length (14-day difference), 
  # the result will be the same as the 15-day fixed-episode example above 
df$r1 <- rolling_episodes(df$date, case_length = 14, recurrence_from_last = F, 
                          to_s4 = T, display = F)

df
```

**Figure 3a: Last event in a window taken as the reference event for the next recurrence window (default)**
```{r message=FALSE, warning=FALSE}
df$r2 <- rolling_episodes(df$date, case_length = 14, recurrence_length = 16, display = F)

df$r2
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df$r2, date = df$date, case_length =14, recurrence_length = 16)
```

**Figure 3b: First event in a window taken as the reference event for the next recurrence window**
```{r message=FALSE, warning=FALSE}
df$r3 <- rolling_episodes(df$date, case_length = 14, recurrence_length = 16, 
                          recurrence_from_last = F, display = F)

df$r3
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df$r3, date = df$date, case_length =14, recurrence_length = 16)
```

### Case window for both case and recurrent events
From `diyar v0.1.0`, you can use the `case_for_recurrence` argument to specify that both `case` and `recurrence events` should have a `case window`. The default behaviour is that only `case events` have a `case window`. See a demonstration of this below. 

*Example 1*

```{r message=FALSE, warning=FALSE}
df <- c("01/04/2019", "11/04/2019", "16/04/2019","21/04/2019","25/04/2019", "07/05/2019")
df <- data.frame(date= as.Date(df, "%d/%m/%Y"))
```

**Figure 4a: Case windows for case events only (default)**
```{r message=FALSE, warning=FALSE}
df$r4 <- rolling_episodes(df$date, case_length = 10, display = F)

df$r4
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df$r4, date = df$date, case_length =10, recurrence_length = 10)
```

**Figure 4b: Case windows for case and recurrence events**
```{r message=FALSE, warning=FALSE}
df$r5 <- rolling_episodes(df$date, case_length = 10, case_for_recurrence = T, 
                          to_s4 = T, display = F)

df$r5
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df$r5, date = df$date, case_length =10, recurrence_length = 10)
```

*Example 2.*

```{r message=FALSE, warning=FALSE}
date <-  seq.Date(as.Date("01/04/2019", "%d/%m/%Y"), 
                  as.Date("21/04/2019","%d/%m/%Y"), 
                  by="3 day")
# Example 2
df2 <- data.frame(date = date, sn = 1:length(date))
# dates
df2$date
```

**Figure 5a: Case windows for case events only (default)**
```{r message=FALSE, warning=FALSE}
df2$r1 <- rolling_episodes(df2$date, case_length = 6, recurrence_length = 4, display = F, sn=df2$sn)

df2$r1
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df2$r1, date = df2$date, case_length =6, recurrence_length = 4)
```

**Figure 5b: Case windows for case and recurrence events**
```{r message=FALSE, warning=FALSE}
df2$r2 <- rolling_episodes(df2$date, case_length = 6, recurrence_length = 4, 
                           case_for_recurrence = T, display = F, sn=df2$sn)

df2$r1
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = df2$r2, date = df2$date, case_length =6, recurrence_length = 4)
```

Notice that in the examples above, all `events` have been assigned to the same `episode` but the number of `recurrent events` differ.

### Number of recurrence periods and episodes
You can choose how many times to check for `episodes` using `episode_max`. When each `strata` of the dataset has the maximum number of episodes specified (`episode_max`), subsequent `events` are given unique episode IDs, making them unique `cases`.

```{r warning=FALSE, message=FALSE}
data("infections_3");
dbs <- infections_3[c("pid","date")]; 
# dates
dbs$date
```

**Figure 6a: Maximum of one fixed episode grouping per strata**
```{r message=FALSE, warning=FALSE}
dbs$eps_1 <- fixed_episodes(strata = dbs$pid, date = dbs$date, case_length = 3, display = F, episodes_max = 1)

dbs$eps_1
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$eps_1, date = dbs$date, case_length =3)
```

**Figure 6b: Maximum of two fixed episode grouping per strata**
```{r message=FALSE, warning=FALSE}
dbs$eps_2 <- fixed_episodes(strata = dbs$pid, date = dbs$date, case_length = 3, display = FALSE, episodes_max = 2)

dbs$eps_2
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$eps_2, date = dbs$date, case_length =3)
```

Similarly, you can choose how many `periods of recurrence` your case definition should permit i.e. how many times to check for `recurrent events`. You do this with `rolls_max`. The default is `Inf` i.e. check until there are no more `events` within the `recurrence period` of the `reference event`. See below.

**Figure 6c: Infinite recurrence periods per episode per strata (default)**
```{r message=FALSE, warning=FALSE}
dbs$eps_4 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = F)

dbs$eps_4
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$eps_4, date = dbs$date, case_length =2, recurrence_length = 2)
```

**Figure 6d: Maximum of one recurrence periods per episode per strata**
```{r message=FALSE, warning=FALSE}
dbs$eps_6 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = F, 
                              rolls_max = 1)

dbs$eps_6
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$eps_6, date = dbs$date, case_length =2, recurrence_length = 2)
```

Note that if `case_for_recurrence` is `TRUE`, after checking for `recurrence` in the last `recurrence period`, the function will also check for the for `duplicates` within the `case window` of the last `recurrent event`. See below.

**Figure 6e: Maximum of one recurrence periods per episode per strata with case_for_recurrence as TRUE**
```{r message=FALSE, warning=FALSE}
dbs$eps_7 <- rolling_episodes(strata = dbs$pid, date =dbs$date, case_length = 2,display = F, 
                              rolls_max = 1, case_for_recurrence = T)

dbs$eps_7
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$eps_7, date = dbs$date, case_length =2, recurrence_length = 2)
```

<a id ="chrono_order"></a>

### Chronological order of events
By default, episode grouping begins at the **earliest** `event` and proceeds to the most recent one, making the earliest `event` the `case`. The opposite of this behaviour is to begin episode grouping at the **most recent** `event` (making that the `case`) and then proceed backwards in time. These behaviours are determined by `from_last`.

```{r message=FALSE, warning=FALSE}
dbs <- infections_2[c("date")]; 
# dates
dbs$date
```

**Figure 7a: Episode grouping in chronological order**
```{r message=FALSE, warning=FALSE}
dbs$forward <- fixed_episodes(date=dbs$date, case_length= 14,
                              group_stats = T, display = F, to_s4=T)

dbs$forward
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$forward, date = dbs$date, case_length = 14)
```

**Figure 7b: Episode grouping in reverse chronological order**
```{r message=FALSE, warning=FALSE}
dbs$backward <- fixed_episodes(date=dbs$date, case_length= 14,
                              group_stats = T, display = F, to_s4=T, from_last=T)

dbs$backward
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = dbs$backward, date = dbs$date, case_length = 14, from_last =T)
```

<a id ="custom_sort"></a>

### User defined case assignment
Aside from the chronological order of `events`, you can use specific `events` as `reference events`. This preference is supplied to `custom_sort`. In `fixed_episodes()` and `rolling_episodes()`, `custom_sort` should be a vector whose values when sorted in ascending order specifies the desired preference. For example, `events` with a `custom_sort` value of `1` will be preferentially taken as the start of an `episode` over those with a value of `2`. This is prioritised over the chronological order of those records. See the example below.

```{r message=FALSE, warning=FALSE}
dates <- c("01/04/2019", "05/04/2019", "07/04/2019")
dates <- as.Date(dates,"%d/%m/%Y")
user_sort <- c(2,1,2)

# dates
dates

# preference determined by from_last 
fixed_episodes(date=dates, case_length=6, display=F, group_stats = T)

# user defined preference via custom sort is prioritised before from_last 
fixed_episodes(date=dates, case_length=6, custom_sort = user_sort, display=F, group_stats = T)

# user defined preference via custom sort is prioritised before from_last. 
# duplicates are flagged from both directions
fixed_episodes(date=dates, case_length=6, custom_sort = user_sort, display=F, 
               bi_direction = T, group_stats = T)

```

In the second example, even though the second record occurred after the first, episode grouping began at the second one. This was because the `custom_sort` value (`user_sort`) for the second record was lower than that of the first.

A consequence of using `custom_sort` this way is that record 1 is not grouped together with record 2 even though it's within the `case_length` of record 2. This is because duplicates are tracked in one direction determined by `from_last`. To track duplicates from both directions use `bi_direction` as shown in the third example.   

Note that examples 1 and 3 results in the same number of `duplicates` (and `episodes`) but different `cases` - records 1 and 2 respectively.

***`custom_sort` will follow `R` standard sort behaviour. For example, a `factor` will sort on its levels not the descriptive label.***

For a practical example, see a case definition where *E. coli* urinary tract infections (UTI) are considered precursors to *E. coli* bloodstream infections (BSI). This means that `episodes` need to be created in such a way that; if there are UTI and BSI records within the same `case_length`, the UTI record will be taken as the `case`. 

```{r message=FALSE, warning=FALSE}
dbs <- diyar::infections_2[c("date","infx")]
dbs$infx <- gsub("E. coli ","",dbs$infx)
dbs$infx[c(2,5)] <- "UTI"

dbs

dbs$epids_1 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx, display = F)

dbs$infx_f <- factor(dbs$infx, levels = c("UTI","BSI"))

dbs$epids_2 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx_f, display = F)

dbs$epids_3 <- fixed_episodes(date=dbs$date, case_length=14, 
                custom_sort = dbs$infx_f, display = F, bi_direction = TRUE)

dbs
```

In `epids_2`, after changing the sort preference using factor levels (`infx_f`), record 1 (*E. coli* BSI)  and record 2 (*E. coli* UTI) are no longer part of the same `episode`. This is because record 2, is now the `reference event` where episode grouping began, and since record 1 occurred before record 2, they are not grouped together. `epids_3` is episode grouping when `bi_direction` is used.

***In `episode_group()`, you can implement `custom_sort` in levels. You'll need to create a column for each level. The column names are then supplied to `custom_sort` in the required order (level)***

<a id="stratified_grouping"></a>

## Stratified episode grouping
Episode grouping can be done separately for different subsets (`strata`) of a dataset. Examples of a `strata` could be patient IDs, pathogen, source of infection or any combination of these. `Episodes` will be limited to each `strata`. `episodes` from different `strata` can have a different `case_lengths` and/or `recurrence_lengths`.

**`record_group()` is useful for creating a group identifier which can then be used as a `strata`. See [record group](record_group.html) for further details.**

As an example, using the `infections` dataset, a case definition may specify the following;

- UTI and BSI records are considered different `episodes` regardless of when they occur
- UTI has a `case_length` of 8 days and `recurrence period` of 31 days
- BSI has a `case_length` of 15 days and no `recurrence period`

In the example below, adding `infection` to the `strata` argument means that `"UTI"` and `"BSI"` records will not be in the same `episode`. Furthermore, the `case_length` and `recurrence_length` for each will be different as specified in the `epi` and `recur` columns respectively 
```{r warning=FALSE}
data(infections)
dbs <- infections[c("date","infection")]
dbs <- dbs[dbs$infection%in% c("UTI","BSI"),]
dbs$epi <- ifelse(dbs$infection=="UTI", 7, 14)
dbs$recur <- ifelse(dbs$infection=="UTI", 30, 0)

dbs$epids <- rolling_episodes(date=dbs$date, case_length =dbs$epi, to_s4 =TRUE,
                            recurrence_length = dbs$recur, display = FALSE, group_stats = TRUE)

dbs
```

<a id="sub_strata"></a>

### Sub-strata
A sub-strata is created when records within a `strata` have different `case_lengths` or `recurrence_lengths`. The case definition below demonstrates how you can take advantage of this.

- UTI and BSI records are different episodes regardless of when they occur. 
- UTIs have a `case_length` of 8 days and recurrence period of 31 days
- BSIs have a `case_length` of 15 days if not treated **OR** a `case_length` of 5 days if treated. In both situations there is no recurrence period.
- Respiratory tract infections (RTI) have a `case_length` of 29 days and recurrence period of 6 days

In this example, whether or not the infection is treated should be viewed as the sub-strata, and the source of infection as the `strata`.

```{r warning=FALSE}
data("infections_4"); 
dbs <- infections_4

dbs$epids <- episode_group(dbs, sn=rid, strata=c(pid, organism, source), date=date, 
                case_length =epi, episode_type = "rolling", recurrence_length = recur,
                display = FALSE)

dbs
```

There are a few things to note with stratified grouping;

- Unless required, `case_length` and `recurrence_length` should be consistent across each `strata` otherwise, you'll inadvertently create a sub-strata
- Episode grouping with and without a sub-strata are different and could lead to different results
- Using a sub-strata is not the same as adding that sub-strata to the `strata` argument. In the example above, adding `treated` to `strata` will group treated infections separately from untreated infections. While this could be the desired outcome depending on your case definition, the case definition above did not required treated and untreated infections to be grouped separately, only that the treated infections last longer.

Stratified grouping is the same as doing a separate analysis for each subset (`strata`) of the dataset.

<a id="interval_grouping"></a>

## Interval grouping
Here we discuss the process of grouping `periods of events` into `episodes`. Each `event period` is essentially a record with a `start_point()` and `end_point()` in time. For the purpose of episode grouping, these `event periods` are created as `number_line()` objects and supplied to the `date` argument. Overlapping `event periods` are then grouped into episodes.

Below is a simple examples;

```{r message=FALSE, warning=FALSE}
dates <- as.Date(c("01/04/2019", "05/04/2019"), "%d/%M/%Y")
dates

# A 10-day period from each date
periods <- number_line(dates, dates + 10)
periods

# Grouping events
fixed_episodes(date=dates, case_length=30, display=FALSE, group_stat=TRUE)

# Grouping periods
fixed_episodes(date=periods, case_length=30, display=FALSE, group_stat=TRUE)
```

For a practical example, below are periods of hospital stay which have been grouped into `episodes`. This is different from grouping the actual admission or discharge events.

```{r message=FALSE, warning=FALSE}
hos_admin <- diyar::hospital_admissions
hos_admin$admin_period <- number_line(hos_admin$admin_dt, hos_admin$discharge_dt)

# Admissions events
hos_admin$admin_dt

# Hospital stay
hos_admin$admin_period
```

**Figure 8a: Grouping the actual admissions into episodes**
```{r message=FALSE, warning=FALSE}
hs_epids_a <- fixed_episodes(date=hos_admin$admin_dt, sn=hos_admin$rd_id, case_length=0, 
                display=F, group_stats=T)
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = hs_epids_a, date = hos_admin$admin_dt, case_length = 0)
```

**Figure 8b: Grouping the periods of stay (admission -> discharge)**
```{r message=FALSE, warning=FALSE}
hs_epids_b <- fixed_episodes(date=hos_admin$admin_period, sn=hos_admin$rd_id, case_length = 0, 
                display=F, group_stats=T)
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = hs_epids_b, date = hos_admin$admin_period, case_length = 0)
```

There are 6 ways `event periods` can overlap (`overlap_methods`). These are `"exact"`, `"across"`, `"inbetween"`, `"chain"`, `"aligns_start"` and `"aligns_end"`. Only `event periods` that overlap by the specified method will be grouped into the same `episode`. You can use any combination of the 6 methods but the default option is to use all 6. See below for a demonstration of this.

```{r warning=FALSE}
# wrapper function to use with lapply
epi_grp_func <- function(x){
  epids <- fixed_episodes(date=hos_admin$admin_period, 
                 sn=hos_admin$rd_id, 
                 case_length = 0, 
                 overlap_methods = x, 
                 display = FALSE, 
                 to_s4 = TRUE, 
                 group_stats = TRUE)
  
  # for the prurpose of this demonstration, limit to overlapped periods    
  epids[epids@epid_total>1]
}

# Methods 
methods <- list(
  # Identical intervals
  exact = "exact",
  # Overlapping intervals
  across= "across",
  # Intervals with aligned start points
  aligns_start = "aligns_start",
  # Intervals with aligned end points
  aligns_end = "aligns_end",
  # Chained intervals
  chain = "chain",
  # Intervals occurring completely within others
  inbetween = "inbetween",
  # Chained intervals and those occurring completely within others
  cb1 = "chain|inbetween",
  # Chained intervals, identical intervals and those occurring completely within others
  cb2 = "exact|chain|inbetween",
  # Chained intervals, overlapping intervals and those with aligned end point
  cb3 = "across|chain|aligns_end"
)

epids <- lapply(methods, epi_grp_func)
names(epids) <- names(methods)

# Results
epids["chain"]

epids["cb2"]
```


**Figure 9a: Grouped episodes when considering only "exact" overlap method**
```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
eps <- epids[["exact"]]
plot_epid(epid = eps, date = hos_admin$admin_period[eps@sn] , case_length = 0)
```

**Figure 9b: Grouped episodes when considering only "chain" overlap method**
```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
eps <- epids[["chain"]]
plot_epid(epid = eps, date = hos_admin$admin_period[eps@sn] , case_length = 0)
```
**Figure 9c: Grouped episodes when considering "chain" or "inbetween" overlap method**
```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
eps <- epids[["cb1"]]
plot_epid(epid = eps, date = hos_admin$admin_period[eps@sn] , case_length = 0)
```

### Case length and recurrence length in interval grouping

From `diyar v0.1.0`, you can use negative `case_lengths` and `recurrence_lengths`. 

+ Using positive "lengths" will collapse two periods into one, if the second one is within* the first one or `n`-days^+^ after the end of the first one. 
+ Using negative "lengths" will collapse two periods into one, if the second one is within* the start and `n`-days^+^ before the end of the first one.
+ Using negative "lengths" that's larger than the duration of the period will collapse the two periods into one, if the second one is within* `(n + 1 + number_line_width())`-days^+^ before the start of the first one.
    + **note that this will happen even if `from_last` is TRUE**

*within or any other method of `overlap`

*^+^ days or any other `episode_unit`*

**Examples below**

```{r warning=FALSE}
pa <- c("28/03/2019", "01/04/2019", "03/04/2019", "07/04/2019","12/04/2019")
pz <- c("31/03/2019", "10/04/2019", "05/04/2019", "09/04/2019","13/04/2019")

pa <- as.Date(pa, "%d/%m/%Y")
pz <- as.Date(pz, "%d/%m/%Y")

periods <- number_line(pa, pz)
periods

```

**Figure 9d: Fixed episodes with a case_length of 6**
```{r message=FALSE, warning=FALSE}
eps_a <- fixed_episodes(periods, case_length =6, group_stats = T, display = F)

eps_a
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = eps_a, date = periods, case_length = 6)
```

Use positive "lengths" when you need to capture periods within a specified duration after the end of a `reference period`. 

**Figure 9e: Fixed episodes with a case_length of -2**
```{r message=FALSE, warning=FALSE}
eps_b <- fixed_episodes(periods, case_length =-2, group_stats = T, display = F)

eps_b
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = eps_b, date = periods, case_length = -2)
```

**Figure 9f: Fixed episodes with a case_length of -6**
```{r message=FALSE, warning=FALSE}
eps_c <- fixed_episodes(periods, case_length =-6, group_stats = T, display = F)

eps_c
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = eps_c, date = periods, case_length = -6)
```

Use negative "lengths" when you only want to capture periods that at least ended `n-days`^+^ before the end of the `reference period`.

It is possible to use a negative "length" that's so large, the `reference period` will capture periods that occur before itself. Consider this a record level implementation of `bi_direction`. Just like `bi_direction`, this is only relevant when you use a `custom_sort`. See below. 

**Figure 9g: Fixed episodes with a case_length of -12**
```{r message=FALSE, warning=FALSE}
custom_sort <- c(2,0,2,1,2)
eps_d <- fixed_episodes(periods, case_length =-12, custom_sort = custom_sort, group_stats = T, display = F)

eps_d
```

```{r echo=FALSE, fig.height=5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_epid(epid = eps_d, date = periods, case_length = -12)
```


<a id="tips"></a>

## Useful ways of using these functions
### Episode grouping across other units of time
In the examples above, episode grouping was done by days (`episode_unit`). However, it can be done by other units of time e.g. hours, days or weeks. Acceptable options are "seconds", "minutes", "hours", "days", "weeks", "months" or "years". Below is an example of episode grouping by the hour.

```{r warning=FALSE, message=FALSE}
data("hourly_data"); dbs <- hourly_data

dbs$datetime

# Each unit is relative to a set number of seconds. 
diyar::episode_unit

# 5-hr episodes
rolling_episodes(date = dbs$datetime, case_length = 5,
                 episode_unit = "hours", recurrence_length = 9, group_stats = TRUE, display = FALSE)

# 5-day episodes
rolling_episodes(date = dbs$datetime, case_length = 5,
                 episode_unit = "days", recurrence_length = 9, group_stats = TRUE, display = FALSE)
```

### Limit episode grouping to a subset of the dataframe
For example, with the `hourly_data` dataset, you can decide to exclude `"GP1"` and `"GP2"` records from episode grouping. Excluded records are skipped and assigned unique episode IDs.

```{r warning=FALSE, message=FALSE}
dbs <- head(hourly_data[c("datetime","category")], 10)
dbs$subset <- ifelse(dbs$category!="GP3", NA, "group")

dbs$epids <- rolling_episodes(strata= dbs$subset, date = dbs$datetime, case_length = 5, episode_unit = "hours", 
                        recurrence_length = 9, display = TRUE, group_stats = TRUE)

dbs
```

### Use a `strata` from `record_group()`
You can used `pid` objects from `record_group()` as a `strata`.

```{r warning=FALSE, message=FALSE}
data(infections) 

dbs <- infections[c("date","infection")]; dbs

# familiar unique record ids for reference - optional
dbs$rd_id <- c(640,17,58,21,130,79,45,300,40,13,31)

# strata based on matching sources of infection
dbs$pids <- record_group(dbs, sn = rd_id,  criteria = infection, display = FALSE)

# stratified grouping 
dbs$epids <- fixed_episodes(sn = dbs$rd_id, date = dbs$date, strata = dbs$pids, 
                             to_s4 = TRUE, display = FALSE, case_length = 10)

dbs
```

### Use `numeric` values for the `date` argument 
`date` and `datetime` objects are fundamentally numeric values within the context of a reference point (`as.Date(..., origin)`). The episode grouping process ignores this context and instead uses each value as a reference point for others. Therefore, it will allow you use `numeric` values instead of `date` or `datetime` values for the `date` argument. For example, if you use `0-2`, it does not matter what they represent i.e. number of days from a particular date or number of seconds from a particular time. It only matters that `0` is before `1`, which in turn is before `2`. When the functions are used this way, the `episode_unit` argument is ignored.

```{r warning=FALSE, message=FALSE}
vals <- c(8.1,6,12,8.5,12,3,8,15,5,7)

vals

fixed_episodes(vals, case_length = .5, group_stats = T, display = F)

fixed_episodes(vals, case_length = 5, group_stats = T, display = F)

fixed_episodes(vals, case_length = 100, group_stats = T, display = F)
```

# Conclusion
There are a variety of ways to use these function. Before beginning, It's worth reviewing your case definition and its implication on the dataset. In general, the following steps will guide you on how to use them;

1. Work out the `strata` in your dataset
2. Choose whether you need `"fixed"` or `"rolling"` episodes
3. Choose whether you are grouping individual events or periods of events by supplying a `date`,`datetime` or `number_line` object as required. See [interval grouping](#interval_grouping)
4. Choose the required `case_length`, and/or `recurrence_length`. The values should be unique to each `strata` unless you require a [sub-strata](#sub_strata)
5. Change `from_last` to `TRUE` if you want to start episode grouping at the most recent record thereby making it the `"case"`. Note that this is not the same as starting episode grouping at the earliest record (`from_last` is `FALSE`) and then picking the most recent record in that episode as the `"case"`. See [chronological order of events](#chrono_order)
6. If you require the `"case"` to be the earliest or most recent event of a particular type of event, use `custom_sort` in combination with `from_last`. If not, ignore this argument. See [user defined case assignment](#custom_sort)
7. If you require episodes to include records on either side of the `"case"` use `bi_direction`. If not, ignore this argument
8. Determine if you are measuring episodes by the minute, hour or day e.t.c., and set `episode_unit` accordingly. Be sure to supply a `datetime` to the `date` argument if `episode_unit` is less than "days"
