---
title: "Number lines and overlaps"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Number lines and overlaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
plot_number_line <- diyar:::plot_number_line
```

#Introduction
`number_line` objects are a range of `numeric` based values. In `diyar`, they represent ranges and periods. How they overlap determine successful matches (see [range matching ](record_group.html#range_matching)) in `record_group()` and overlapping windows (see [interval grouping](episode_group.html#interval_grouping)) in `fixed_episodes()`, `rolling_episodes()` and `episode_group()`. 

#Overview
```{r include=FALSE}
library(diyar)
exp <- number_line(8, 1)

```

For a given `number_line` object e.g. ``r format(exp)``, its main components are;

+ `left_point()` - left side of the `number_line` `i.e.` ``r left_point(exp) `` 
+ `right_point()` - right side of the `number_line` `i.e.` ``r right_point(exp) ``
+ `start_point()` - lower end of the `number_line` `i.e.` ``r start_point(exp) ``
+ `end_point()` - upper end of the `number_line` `i.e.` ``r end_point(exp) ``
+ `direction` - `"decreasing"` `(<-)`, `"increasing"` `(->)` or no direction `(==)`
+ `number_line_width()` - difference between the left and right points `i.e.` ``r number_line_width(exp) ``

In `"increasing"` `number_line` objects, the `start_point()` is on the `left_point()` while in `"decreasing"` `number_line` objects, the `start_point()` is on the `right_point()`.

```{r message=FALSE, warning=FALSE}
library(diyar)

nl_a <- number_line(l=c(3,5), r = c(4,1))

# `number_line` objects
nl_a

# lower end of the range
start_point(nl_a)

# start of the range
left_point(nl_a)
```

**Figure 1: `number_line` objects **
```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(nl_a)
```

You can reverse the direction of a `number_line` object with `reverse_number_line()`.

```{r message=FALSE, warning=FALSE}
# reverse only decreasing `number_line` objects
nl_b <- reverse_number_line(nl_a, direction = "decreasing"); nl_b
```

**Figure 2: Direction of `number_line` objects **
```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(nl_b)
```

## Manipulating `number_line` objects
There are convenience functions to reverse, shrink or expand `number_line` objects, or shift them along a the number line. 

```{r message=FALSE, warning=FALSE}
nl_c <- number_line(as.Date("04/04/2019", "%d/%M/%Y"),
                    as.Date("07/04/2019", "%d/%M/%Y"))

nl_c

# expand the `number_line` object by 2 days from both ends 
nl_d <- expand_number_line(nl_c, 2); nl_d

# expand the `number_line` object by 2 days from the lower end
nl_e <- expand_number_line(nl_c, 2, "start"); nl_e

# shrink the `number_line` object by 2 days from the upper end
nl_f <- expand_number_line(nl_c, -2, "end"); nl_f

# shrink the `number_line` object by 2 days from both ends
nl_g <- expand_number_line(nl_c, -2); nl_g

# reverse the direction of the `number_line`
nl_h <- reverse_number_line(nl_c); nl_h

# shift the `number_line` object towards the left of the number line by 2 days 
nl_g <- shift_number_line(nl_c, -2); nl_g

# shift the `number_line` object towards the right of the number line by 2 days 
nl_i <- shift_number_line(nl_c, 2); nl_i
```

**Figure 3: Manipulated `number_line` objects **
```{r echo=FALSE, fig.height=2.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(c(nl_c, nl_d, nl_e, nl_f, nl_g, nl_h, nl_i))
```

You can also convert `number_line` objects to a sequence of `numeric` based values

```{r message=FALSE, warning=FALSE}
nls <- c(nl_c, nl_d, nl_e, nl_f, nl_g, nl_h, nl_i)

nls

number_line_sequence(nls, by=2)
```

## Overlaps
`number_line` objects can overlap with each other and do so in six ways - `"exact"`, `"inbetween"`, `"across"`, `"chain"`, `"aligns_start"` and `"aligns_end"`. Each pair of `number_line` objects can only overlap by one these methods. For example, a `numberline` object can not have aligned start and end points, instead it's considered an `"exact"` overlap.

**Figure 4a: Example of `exact` overlaps**

```{r message=FALSE, warning=FALSE}
exact <- c(number_line(1,2), number_line(1,2),
           # negative test - start_point() and end_point() 
           # for both must be identical to be an exact mactch
           number_line(3,4), number_line(4,3), 
           number_line(2.5,2.5), number_line(2.5,2.5))

# positive logical test
exact(exact[c(1,3,5)], exact[c(2,4,6)])
# negtaive logical test
across(exact[c(1,3,5)], exact[c(2,4,6)])
# check overlap methods as defined in `diyar`
overlap_method(exact[c(1,3,5)], exact[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(exact, show_overlap=T)
```

**Figure 4b: Example of `inbetween` overlaps**

```{r message=FALSE, warning=FALSE}
inbetween <- c(number_line(5, 8), number_line(6, 7), 
               number_line(11, 10), number_line(9, 12), 
               number_line(13, 15), as.number_line(14))

# positive logical test
inbetween(inbetween[c(1,3,5)], inbetween[c(2,4,6)])
# negtaive logical test
across(inbetween[c(1,3,5)], inbetween[c(2,4,6)])
# check overlap methods as defined in `diyar`
overlap_method(inbetween[c(1,3,5)], inbetween[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(inbetween, show_overlap=T)
```

**Figure 4c: Example of `across` overlaps **

```{r message=FALSE, warning=FALSE}
across <- c(number_line(16, 19), number_line(18, 21), 
            number_line(22, 25), number_line(23, 26),
            number_line(27, 30), number_line(31, 29))

# positive logical test
across(across[c(1,3,5)], across[c(2,4,6)])
# negtaive logical test
inbetween(across[c(1,3,5)], across[c(2,4,5)])
# check overlap methods as defined in `diyar`
overlap_method(across[c(1,3,5)], across[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(across, show_overlap=T)
```

**Figure 4d: Example of `chain` overlaps**

```{r message=FALSE, warning=FALSE}
chain <- c(number_line(32, 34), number_line(34, 36), 
            number_line(39, 37), number_line(41, 39), 
           # negative test - end_point() of one must lead to the start_point() 
           # to be considered a chain overlap
           number_line(42, 44), number_line(47, 44))

# positive logical test
chain(chain[c(1,3,5)], chain[c(2,4,6)])
# negtaive logical test
across(chain[c(1,3,5)], chain[c(2,4,6)])
# check overlap methods as defined in `diyar`
overlap_method(chain[c(1,3,5)], chain[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(chain, show_overlap=T)
```


**Figure 4e: Example of `aligns_start` overlaps**

```{r message=FALSE, warning=FALSE}
aligns_start <- c(number_line(45, 46), number_line(45, 47), 
            number_line(48, 49.5), number_line(48, 48),
            number_line(51, 50), number_line(51, 52))

# positive logical test
aligns_start(aligns_start[c(1,3,5)], aligns_start[c(2,4,6)])
# negtaive logical test
across(aligns_start[c(1,3,5)], aligns_start[c(2,4,6)])
# check overlap methods as defined in `diyar`
overlap_method(aligns_start[c(1,3,5)], aligns_start[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(aligns_start, show_overlap=T)
```

**Figure 4f: Example of `aligns_end` overlaps**

```{r message=FALSE, warning=FALSE}
aligns_end <- c(number_line(54, 55), number_line(53, 55), 
            number_line(56, 57.5), number_line(57.5, 57.5), 
            number_line(58, 59), number_line(60, 59))

# positive logical test
aligns_end(aligns_end[c(1,3,5)], aligns_end[c(2,4,6)])
# negtaive logical test
across(aligns_end[c(1,3,5)], aligns_end[c(2,4,6)])
# check overlap methods as defined in `diyar`
overlap_method(aligns_end[c(1,3,5)], aligns_end[c(2,4,6)])
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(aligns_end, show_overlap=T)
```

## Compressing `number_line` objects
You can compress a pair of overlapping `number_line` objects into one using `compress_number_line()`. You can also choose to compress only those that overlap in a certain way among the 6 methods shown above.

**Figure 5a: Pairs of `number_line` objects overlapping by `"inbetween"` and `"exact"` methods**
```{r message=FALSE, warning=FALSE}
cnl_a <- c(inbetween, exact)
cnl_a
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(c(inbetween, exact), show_overlap=T)
```

**Figure 5b. Pairs of `number_line` objects overlapping by `"inbetween"` and `"exact"` methods compressed into one**
```{r message=FALSE, warning=FALSE}
cmp_a <- compress_number_line(cnl_a, deduplicate = T)
cmp_a
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
f <- number_line(left_point(c(cnl_a, cmp_a)), right_point(c(cnl_a, cmp_a)), gid =c(cnl_a@gid, cmp_a@gid), id =c(cnl_a@id, cmp_a@id)  )
plot_number_line(f, show_overlap=F)
```

**Figure 6a: Pairs of `number_line` objects overlapping by `"across"` and `"chain"` methods**
```{r message=FALSE, warning=FALSE}
cnl_b <- c(across, chain)
cnl_b
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(c(across, chain), show_overlap=T)
```

**Figure 6b. Pairs of `number_line` objects overlapping by `"across"` and `"chain"` methods compressed into one**
```{r message=FALSE, warning=FALSE}
cmp_b <- compress_number_line(cnl_b, deduplicate = T)
cmp_b
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
f <- number_line(left_point(c(cnl_b, cmp_b)), right_point(c(cnl_b, cmp_b)), gid =c(cnl_b@gid, cmp_b@gid), id =c(cnl_b@id, cmp_b@id)  )
plot_number_line(f, show_overlap=F)
```

**Figure 7a: Pairs of `number_line` objects overlapping by `"aligns_start"` and `"aligns_end"` methods**
```{r message=FALSE, warning=FALSE}
cnl_c <- c(aligns_start, aligns_end)
cnl_c
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(c(aligns_start, aligns_end), show_overlap=T)
```

You can choose to only compress `number_line` objects that overlap in a particular way using the `methods` argument.

**Figure 7b. Pairs of `number_line` objects overlapping by `"aligns_start"` and `"exact"` methods compressed into one**
```{r message=FALSE, warning=FALSE}
cmp_c <- compress_number_line(cnl_c, deduplicate = T, methods ="aligns_start|exact")
cmp_c
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
f <- number_line(left_point(c(cnl_c, cmp_c)), right_point(c(cnl_c, cmp_c)), gid =c(cnl_c@gid, cmp_c@gid), id =c(cnl_c@id, cmp_c@id))
plot_number_line(f, show_overlap=F)
```

## Collapsing `number_line` objects
The resulting `number_line` objects from a compressed pair can yet again overlap with others, especially when not all `methods` were used. By default, resulting compressed `number_line` objects are not compressed a second time however, you can do so using the `collapse` argument. It's worth mentioning that this process starts from the 1st `number_line` object and then proceeds to the last in the set. Therefore, their position in the set can lead to different results. 

In the figure below, there are `"across"`, `"chain"`, `"inbetween"` and `"aligns_end"` overlaps.

**Figure 8a: Pairs of `number_line` objects overlapping by `"across"`, `"chain"`, `"inbetween"` and `"aligns_end"` methods**
```{r message=FALSE, warning=FALSE}
nl_z <- number_line(l= c(1,2,4,5,6), r = c(3,4,8,8,7))
nl_z
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(nl_z, show_overlap=T)
```

In the example below, we chose to only compress the pairs that overlap by `"chain"` and `"aligns_end"` methods. The resulting `number_line` object from compressing ``r paste0(format(nl_z[1:2]), collapse=" and ")`` now overlaps with the resulting `number_line` objects from compressing ``r  paste0(format(nl_z[3:4]), collapse=" and ")``.
 
**Figure 8b: Pairs of `number_line` objects overlapping by `"chain"` and `"aligns_end"` methods compressed into one**

```{r message=FALSE, warning=FALSE}
cmp_z <- compress_number_line(nl_z, methods = "across|chain")

cmp_z
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(cmp_z, show_overlap=T)
```

As mentioned, if each `number_line` object's position in the set changes, the result could also change. See below. 

**Figure 8c: Rearranged pairs of `number_line` objects overlapping by `"across"`, `"chain"`, `"inbetween"` and `"aligns_end"` methods **

```{r message=FALSE, warning=FALSE}
nl_zb <- nl_z[c(2:5,1)]; nl_zb

cmp_zb <- compress_number_line(nl_zb, methods = "across|chain"); cmp_zb
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(cmp_zb, show_overlap=T)
```

In Fig. 8b, ``r format(nl_z[1])`` came first, so only `number_line` objects that overlapped with ``r format(nl_z[1])`` were compressed. However in Fig. 8c, because of the new positions, ``r format(nl_zb[1])`` came first, so only those that overlapped with ``r format(nl_z[1])`` were compressed. 

However, you can use the `collpase` argument to further compress already compressed `number_line` objects if they overlap again. In the example below, we collapse (rather than compress) the initial set (`nl_zb`) which gives the same result as compressing the rearranged version. 

**Figure 8d: pairs of `number_line` objects overlapping by `"chain"` and `"aligns_end" methods collapsed into one**

```{r message=FALSE, warning=FALSE}
cmp_zba <- compress_number_line(nl_z, methods = "across|chain", collapse = T); cmp_zba
```

```{r echo=FALSE, fig.height=1.5, fig.width=8.5, message=FALSE, warning=FALSE}
plot_number_line(cmp_zba, show_overlap=T)
```

\newline

`fixed_episodes()`, `rolling_episodes()` and `episode_group()` are more advanced implementations of `compress_number_line()`. They can compress (`"fixed_episodes"`) and collapse (`"rolling_episodes"`) overlapping `number_line` objects. They have additional useful features such as controlling the sequence in which pairs are compressed i.e. chronological order (`from_last`) or a user defined order (`custom_sort`), and considering add periods (`case_length` and `recurrence_length`) before and/or (`bi_direction`) after the reference `number_line` object. See [episode grouping](episode_group.html) for further details.

