---
title: "episodes_wf_splits"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reduce the number of records being assessed
Reframing the dataset in such a way that you have to deal with fewer records can also save processing time. For example, assume a dataset of 10 events from 5 individuals on 3 separate days, and a case definition to assess a potential transmission event between these individuals. Instead of analysing 10 events (records) with the person identifier as a sub-criteria, change the dataset to 3 dates (records) with a `list` of all individuals per date. The `list` can then be passed to `case_sub_criteria` or `recurrence_sub_criteria` for the necessary logical test(s). `episodes_wf_splits()` is a wrapper function of `episodes()` designed to achieve this. It is particularly useful when a `sub_criteria` is used. Below is an example of the used case described above.

```{r warning = FALSE}
# One event every day of the year
event_dt2 <- seq(from = as.Date("2020-01-01"), to = as.Date("2020-01-31"), by = 1)
# The same event reported by 100 individuals on the same day
dup_events_b <- rep(event_dt2, 2)
# Attribute 1 - 36,600 unique individuals
attr_4 <- seq_len(length(dup_events_b))
attr_4 <- rep(1, length(dup_events_b))
# Fast approach
  # Logical test - 2 or more different individuals within a `case_length`
  match_funx_a <- function(x, y){
    splts <- split(x$person, y$sn)
    sn_sets <- as.numeric(names(splts))
    splts <- lapply(splts, function(x) length(x[!duplicated(x)]) > 1)
    splts <- unlist(splts, use.names = FALSE)
    lgk <- splts[match(y$sn, sn_sets)]
    lgk
  }
  # Equivalence of records. Still required
  equal_funx <- function(x, y) FALSE
  # Create the `sub_criteria`
  sub_cri_7a <- sub_criteria(list(person = attr_4,
                               sn = 1:length(attr_4)), 
                          match_funcs = match_funx_a, 
                          equal_funcs = equal_funx)
system.time(
  ep24a <- episodes(dup_events_b, case_length = 1, 
                    case_sub_criteria = sub_cri_7a)  
)

# Faster approach
  date_strata_cmbi <- date_strata_combi(strata = rep(1, length(dup_events_b)), 
                                        date = dup_events_b)
  # Reframed attribute 1 - List of unqiue 100 patients per day 366 days
  attr_4b <- split(attr_4, date_strata_cmbi)
  attr_4b <- list(persons = attr_4b, 
                  sn = 1:length(attr_4b))

  # Reframed logical test - 2 or more different individuals within a `case_length`
  match_funx_b <- function(x, y){
    splts <- split(x$persons, y$sn)
    sn_sets <- as.numeric(names(splts))
    splts <- lapply(splts, function(x) {
      x <- unlist(x, use.names = FALSE)
      length(x[!duplicated(x)]) > 1
    })
    splts <- unlist(splts, use.names = FALSE)
    lgk <- splts[match(y$sn, sn_sets)]
    lgk
  }
  # Create the `sub_criteria`
  sub_cri_7b <- sub_criteria(attr_4b, 
                            match_funcs = match_funx_b, 
                            equal_funcs = equal_funx)
# Faster
system.time(
  ep24b <- episodes_wf_splits(dup_events_b, case_length = 1,
                              case_sub_criteria = sub_cri_7b)  
)

# Same outcomes - identical identifiers
all(ep24a == ep24b)

# and same number of iterations
max(ep24a@iteration); max(ep24b@iteration)
```
